Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Final State with Multivariate Techniques

### 2015-12-04 paper/sec-pileup.tex
more work on the paper

The main limitation of the analysis of the previous section was that
pile-up (PU) effects were
not accounted for.
%
However, any realistic feasibility study at the HL-LHC, specially
like the present one that relies on the detailed modeling
of kinematic distributions, needs to assess the robustness of the conclusions
with respect to PU.
%
In this section, we show how our qualitative results, in particular
the enhanced signal significance provided by the MVA, are robust
in the presence of realistic PU conditions.
%
First of all we describe the simulation of realistic
PU conditions at the HL-LHC, and discuss the settings of
the PU subtraction strategy that we adopt.
%
Then we present the validation of the PU subtraction,
and compare a number of distributions, including substructure variables,
with and without PU.
%
Finally we revisit the MVA analysis of Sect.~\ref{sec:mva}, and
show that our qualitative conclusions are robust
in the presence of realistic PU conditions, finding
a combined
signal significance of $S/\sqrt{B}\simeq 4$.

@@

To study the impact of PU in our analysis,
Minimum Bias (MB) events,
including Multiple Parton Interactions (MPI), have been generated
with {\tt Pythia8}, and
superimposed to the signal
and background samples described in Sect.~\ref{mcgeneration}.
%
We have explored two scenarios for the amount of PU expected
at the HL-LHC, one with a mean number of
PU vertices per bunch crossing of $\la n_{\rm PU}\ra=80$, and another
with $\la n_{\rm PU}\ra=150$.
%
In order to subtract the PU, a number of techniques
have been developed
recently~\cite{Cacciari:2009dp,TheATLAScollaboration:2013pia,Butterworth:2008iy,Cacciari:2007fd,Krohn:2009th,Krohn:2013lba,Ellis:2009me,Bertolini:2014bba,Cacciari:2014gra,Cacciari:2014jta,Berta:2014eza,Larkoski:2014wba}.
%
In this work, PU  will be subtracted by means
of the the {\tt SoftKiller} (SK)
method~\cite{Cacciari:2014gra}, as implemented in {\tt FastJet}.
%

The idea underlying {\tt SoftKiller} is based on eliminating particles
below a given cut-off in their transverse momentum, $p_T^{\rm (cut)}$, whose
value is dynamically determined in a way that makes the event-wide
transverse-momentum flow density $\rho$ vanish.
%
This $p_T$ flow density is defined as
\be
\rho\equiv{\rm median}_i \Bigg\{ \frac{p_{Ti}}{A_i}\Bigg\} \, ,
\ee
where the median is computed over all the patches $i$ with area
$A_i$ and transverse momentum $p_{Ti}$ in which the $\lp \eta,\phi\rp$ plane
is partitioned.
%
From its definition in terms of the median,
we observe that the value of $p_T^{(\rm cut)}$
will be dynamically raised until half of the patches have $\rho=0$.
%
The size and number of these patches is a free parameter of the algorithm -
here we will use square patched with length $a=0.4$.
%
We restrict ourselves to the central rapidity interval,
$\eta \in \lc -2.5, 2.5\rc$, for the estimation of the
$p_T$ flow density $\rho$.
%
In this analysis, {\tt SoftKiller} method is applied
to particles at the end of the parton shower, before
the jet clustering stage.

%
We now compare different kinematical distributions
in the three relevant cases:
cases:
\begin{itemize}
\item without any PU,
\item adding PU, but without any PU subtraction,
  \item adding PU, followed by the {\tt SoftKiller} subtraction.
\end{itemize}
As we will show now, {\tt SoftKiller} exhibits a reasonable
performance in PU subtraction.
%
We will also see that the boosted category is less affected
by PU than the resolved category, as expected from the higher
$p_T$ threshold in the corresponding selection cuts.

Let us begin by comparing
a number of signal distributions with
and without PU, with $\la n_{\rm PU}\ra=80$ in the former
case.
%
In Fig.~\ref{fig:m_H_PU} we show the invariant mass distribution
of the leading and subleading Higgs candidates, corresponding to both
the boosted
and resolved categories.
%
These distributions are plotted after the $b$-tagging, that is,
before they are used as input to the MVA.
%
As we can see, in the boosted category, the residual effects of PU
after the {\tt SoftKiller} subtraction are very mild in these
distributions, with the position of the Higgs mass peaks essentially
unchanged.
%
The effects are more important for the resolved category, where PU
shifts the Higgs peak by an amount $\Delta m_h \simeq 5$ GeV.

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
  \begin{center}
      \vspace{-1cm}
      \includegraphics[width=0.49\textwidth]{plots/m_H0_res_comp.pdf}
      \includegraphics[width=0.49\textwidth]{plots/m_H1_res_comp.pdf}
      \includegraphics[width=0.49\textwidth]{plots/m_H0_bst_comp.pdf}
      \includegraphics[width=0.49\textwidth]{plots/m_H1_bst_comp.pdf}
  \caption{\small
    Comparison of the invariant mass distributions of the leading (left plots)
    and subleading (right plots) Higgs candidates in the resolved
    (upper plots) and boosted (lower plots) categories,
    without PU and with $\la n_{PU}\ra=80$ subtracted with {\tt SoftKiller}.
}
\label{fig:m_H_PU}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

Next we compare the transverse momentum of the leading Higgs
candidate, $p_t^{h_1}$ and the invariant mass of the di-Higgs system
$m_{hh}$, in Fig.~\ref{fig:mHH_PU}, both for the boosted and
for the resolved categories.
%
In the case of the $p_T$ distribution, the differences between the selection
criteria for the resolved
and boosted categories is reflected in the rightward shift of the latter.
%
The effect of PU is negligible in the boosted case, and small
in the resolved case, except for large $p_T$ values.
%
For the case of the $m_{hh}$ distribution, similar conclusions
apply.
%
Note that this latter observable is particularly sensitive
of BSM dynamics.
%
These results validate the PU subtraction strategy adopted in this work.

@@

Next, we assess the impact of PU in representative
substructure variables
used as input to the MVA in the boosted category.
%
In particular we consider the subjetiness variable,
$\tau_{21}$, and the ratio
of energy correlation functions, $D_2^{(\beta)}$,
corresponding to the the leading Higgs candidate.
%
This comparison is illustrated in Fig.~\ref{fig:Substructure_PU}.
%
As can be seen, these substructure variables, which
as demonstrated before carry a very substantial
discrimination power, are relatively unaffected by PU.
%
Therefore, we do not expect a substantial loss of discrimination
power due to PU effects in our analysis.
%
Note in particular that $D_2^{(\beta)}$ was constructed to be resilient
with respect to PU contamination.

@@

%
We now assess how the signal over background discrimination is affected
in the presence of PU.
%
First of all let us consider the boosted category.
%
In Fig.~\ref{fig:signal-vs-back-boosted} we compare
various kinematical distributions for signal and background events,
     with and without PU: the invariant mass and the $p_T$ of the leading
     Higgs candidate, and the $\tau_{21}$ and $D^{(\beta)}_2$ substructure variables.
     %
     All these distributions correspond to the boosted category.
     %
     While for signal distributions the effects of PU after SK subtraction are small,
     differences are more marked for the background distributions, where
     the shapes of the distributions with and without background can be substantially
     different.
     %
     We note however that the fact that signal and background distribution exhibit
     rather different behaviors is maintained in the presence of PU.

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
  \begin{center}
  \includegraphics[width=0.49\textwidth]{plots/m_h0_bst_comp_back.pdf}
  \includegraphics[width=0.49\textwidth]{plots/pt_h0_bst_comp_back.pdf}
   \includegraphics[width=0.49\textwidth]{plots/tau21_h1_bst_comp_back.pdf}
  \includegraphics[width=0.49\textwidth]{plots/D2_h0_bst_comp_back.pdf}
   \caption{\small
     Comparison of kinematical distributions for signal and background events,
     with and without PU: the invariant mass and the $p_T$ of the leading
     Higgs candidate, and the $\tau_{21}$ and $D^{(\beta)}_2$ substructure variables.
     %
     All these distributions correspond to the boosted category.
}
\label{fig:signal-vs-back-boosted}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

>>>>

The main limitation of our analysis so far is the fact that
pile-up (PU) effects have not been accounted for
not accounted for.
%
However, any realistic feasibility study at the HL-LHC, specially
one 
such the present that relies on the careful modeling
of kinematic distributions,
requires assessing the robustness of the conclusions
with respect to the inclusion of PU.
%
In this section, we show how our qualitative results, in particular
the enhanced signal significance provided by the MVA, are robust
in the presence of  PU.

First of all, we describe the simulation of the 
PU conditions  expected at the HL-LHC, and discuss the settings of
the PU subtraction strategy that we adopt.
%
Then we present the validation of the PU subtraction,
and compare a number of distributions, including substructure variables,
with and without PU.
%
Finally we revisit the MVA analysis of Sect.~\ref{sec:mva}, and
show that our qualitative conclusions are robust
in the presence of realistic PU conditions, finding
a combined
signal significance of $S/\sqrt{B}\simeq 4$.

@@

To study the impact of PU in our analysis,
Minimum Bias (MB) events,
including Multiple Parton Interactions (MPI), have been generated
with {\tt Pythia8}, and
superimposed to the signal
and background samples described in Sect.~\ref{mcgeneration}.
%
We have explored two scenarios for the amount of PU expected
at the HL-LHC, one with a mean number of
PU vertices per bunch crossing of $\la n_{\rm PU}\ra=80$, and another
with $\la n_{\rm PU}\ra=150$.
%
In order to subtract the PU, a number of techniques
have been developed
recently~\cite{Cacciari:2009dp,TheATLAScollaboration:2013pia,Butterworth:2008iy,Cacciari:2007fd,Krohn:2009th,Krohn:2013lba,Ellis:2009me,Bertolini:2014bba,Cacciari:2014gra,Cacciari:2014jta,Berta:2014eza,Larkoski:2014wba}.\footnote{
These techniques have also important applications in the subtraction
of the UE/MPI contamination for jet reconstruction
in heavy ion collisions~\cite{Cacciari:2010te}.
}
%
In this work, PU  will be subtracted by means
of the the {\tt SoftKiller} (SK)
method~\cite{Cacciari:2014gra}, as implemented in {\tt FastJet}.
%

The idea underlying {\tt SoftKiller} is based on eliminating particles
below a given cut-off in their transverse momentum, $p_T^{\rm (cut)}$, whose
value is dynamically determined in a way that makes the event-wide
transverse-momentum flow density $\rho$ vanish.
%
This $p_T$ flow density is defined as
\be
\rho\equiv{\rm median}_i \Bigg\{ \frac{p_{Ti}}{A_i}\Bigg\} \, ,
\ee
where the median is computed over all the patches $i$ with area
$A_i$ and transverse momentum $p_{Ti}$ in which the $\lp \eta,\phi\rp$ plane
is partitioned.
%
From its definition in terms of the median,
we observe that the value of $p_T^{(\rm cut)}$
will be dynamically raised until half of the patches have $\rho=0$.
%
The size and number of these patches is a free parameter of the algorithm -
here we will use square patches with length $a=0.4$.
%
We restrict ourselves to the central rapidity region,
$|\eta| \le 2.5$, for the estimation of the
$p_T$ flow density $\rho$.
%
In this analysis, {\tt SoftKiller} method is applied
to particles at the end of the parton shower, before
jet clustering.

To validate PU subtraction,
we now compare different kinematical distributions
in the case without PU and in the case
with PU subtracted with {\tt SoftKiller}.
%
Then we will also compare some specific distributions
also in the case where PU has not been subtracted.
%
We find that
{\tt SoftKiller} exhibits a reasonable
performance for the PU subtraction.
%
We  also obtain that the boosted category is less affected
by PU than the resolved category, as expected from the higher
$p_T$ threshold in the corresponding selection cuts.

Let us begin by comparing
representative signal distributions with
and without PU, with $\la n_{\rm PU}\ra=80$ in the
former
case (and SK subtraction).
%
In Fig.~\ref{fig:m_H_PU} we show the invariant mass distribution
of the leading and subleading Higgs candidates, corresponding to both
the boosted
and resolved categories.
%
These distributions are plotted after the $b$-tagging, that is,
before they are used as input to the MVA.
%
As we can see, in the boosted category, the residual effects of PU
after the {\tt SoftKiller} subtraction are rather mild,
with the position of the Higgs mass peaks essentially
unchanged, and only a moderate distortion of the
distribution found.
%
The effects are more important for the resolved category, where PU
shifts the Higgs peak by an amount $\Delta m_h \simeq 5$ GeV.

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
  \begin{center}
      \vspace{-1cm}
      \includegraphics[width=0.49\textwidth]{plots/m_H0_res_comp.pdf}
      \includegraphics[width=0.49\textwidth]{plots/m_H1_res_comp.pdf}
      \includegraphics[width=0.49\textwidth]{plots/m_H0_bst_comp.pdf}
      \includegraphics[width=0.49\textwidth]{plots/m_H1_bst_comp.pdf}
  \caption{\small
    Comparison of the invariant mass distributions of the leading (left plots)
    and subleading (right plots) Higgs candidates in the resolved
    (upper plots) and boosted (lower plots) categories,
    both without PU and with
    PU, $\la n_{PU}\ra=80$, subtracted with {\tt SoftKiller}.
}
\label{fig:m_H_PU}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

Next we compare the transverse momentum of the leading Higgs
candidate, $p_t^{h_1}$ and the invariant mass of the di-Higgs system
$m_{hh}$, in Fig.~\ref{fig:mHH_PU}, both for the boosted and
for the resolved categories.
%
In the case of the $p_T$ distribution, the differences between the selection
criteria for the resolved
and boosted categories is reflected in the rightward shift of the latter.
%
The effect of PU is negligible in the boosted case, and small
in the resolved case, except for large $p_T$ values.
%
For the case of the $m_{hh}$ distribution, similar conclusions
apply.
%
These comparisons validate the PU subtraction strategy
adopted in this work.

@@

We can also assess the impact of PU in representative
substructure variables
used as input to the MVA in the boosted category.
%
In particular we consider the subjetiness variable,
$\tau_{21}$, Eq.~(\ref{eq:tau21}), and the ratio
of energy correlation functions, $D_2^{(\beta)}$,
Eq.~(\ref{eq:d2}),
corresponding to the leading Higgs candidate.
%
This comparison is illustrated in Fig.~\ref{fig:Substructure_PU}.
%
As can be seen, these substructure variables, which
as demonstrated before carry a substantial
discrimination power, are relatively unaffected by PU.
%
Recall that the $D_2^{(\beta)}$ variable is
explicitly constructed~\cite{Larkoski:2013eya}
to be resilient
with respect to PU contamination.
%
Therefore, we do not expect a major loss of discrimination
power due to PU effects in our analysis.
%

@@

It is also interesting to quantify how
the relative differences between
signal over background distributions are affected in 
the presence of PU.
%
Considering first of all the boosted category,
in Fig.~\ref{fig:signal-vs-back-boosted} we compare
various kinematical distributions for signal and background events,
     with and without PU: the invariant mass and the $p_T$ of the leading
     Higgs candidate, the 2-jettines $\tau_{21}$ and $\sqrt{d_{12}}$,
     $k_T$ splitting scale, Eq.~(\ref{eq:ktsplitting}),
     for the leading Higgs candidate.
     %
     While for signal distributions the effects of PU after SK subtraction are small,
     differences are more marked for the background distributions, where
     the shapes of the distributions with and without background can be substantially
     different.
     %
     We note however that the fact that signal and background distribution exhibit
     rather different behaviors is maintained in the presence of PU.

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
  \begin{center}
  \includegraphics[width=0.49\textwidth]{plots/m_h0_bst_comp_back.pdf}
  \includegraphics[width=0.49\textwidth]{plots/pt_h0_bst_comp_back.pdf}
   \includegraphics[width=0.49\textwidth]{plots/tau21_h1_bst_comp_back.pdf}
  \includegraphics[width=0.49\textwidth]{plots/split12_h0_bst_comp_back.pdf}
   \caption{\small
     Comparison of kinematical distributions, in
     the boosted category, for signal and background events
     with and without PU: the invariant mass,  $p_T$,
     and the substructure variables $\tau_{21}$ and $\sqrt{d_{12}}$
    for the leading Higgs candidate.
     %
 }
\label{fig:signal-vs-back-boosted}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
